Collaborative Visual Analysis on the Web with RCloud

### 2015-06-25 interviews.tex
fixups

To evaluate the effectiveness of RCloud, we interviewed 13 current or
recent users of RCloud. 9 are data analysts, and 4 build tools for data
analysts and business needs.

The subjects differed in their evaluation of its benefits and disadvantages,
as described below.

@@

Besides providing a way to present work, notebook sharing can provide a
starting point for coding. Lilo says, ``The best part is how easily you can
share code. You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Wendy
notes, ``[If] some person has done something similar, then you're able to just
edit that, and that's saved a lot of work for me.''

Eric develops packages for analysts, and uses RCloud ``more for sharing code with
other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code... I write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets [and display the results].'' He also uses RCloud for
describing and debugging problems with data sources.

@@

Automatic source control is also a popular feature. Lilo says, ``Instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

@@

\subsection{Discovering others' work}
Many users find the search function valuable. Wendy says, ``The fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''

But other users prefer to browse just the notebooks of experts they
know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
through, because I know that the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''

More selective ways to search will be needed as the number of notebooks
grows further. We mention possible approaches in Section~\ref{sec:discussion}.

@@

Integrating R also helps Allison's application: ``Having an open session where I can run R commands or
functions without having to invoke an API or send a request and then
wait for the response is extremely helpful in writing the application.''

@@

Although most of the analysts appreciate RCloud's sharing features,
it is not popular as a tool for exploratory data analysis.
All analysts with R experience prefer to do analysis in another
tool, then paste code into RCloud for sharing.
It appears this is mainly due to familiarity.
Eric asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

\subsection{Cells versus the command line}
RCloud's notebook interface combines editable Markdown with a command line
interface.

The interface takes some learning. Lilo says,
\begin{quote}
%% It took me a couple of weeks of
%% looking at it to become comfortable because 
I saw the cells and it just threw me
for a total loop. I mean, it's a good idea because [...] I can run it all at
once if I want or I can break these into sections for either debugging or
staging purposes. I really like it now, but when I first saw it, [it was] very
confusing.
\end{quote}

@@

The cell structure also can be problematic if some cells take a long time to
execute. High variance is common: cells may take anywhere from seconds to
hours to run. In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''
To avoid this pitfall, some users write code within cells to explicitly cache results.
Coby ends up ``littering the notebook with little switches that comment out the
slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.

>>>>

To evaluate the effectiveness of RCloud, we interviewed 13 current or
recent users of RCloud. 9 are data analysts, and 4 build tools for data
analysts and business needs.
% no paragraphbreak needed! scn
The subjects differed in their evaluation of its benefits and disadvantages,
as described below.

@@

Besides providing a way to present work, notebook sharing become a starting point
in coding. Lilo says, ``The best part is how easily you can
share code. You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Wendy
notes, ``[If] some person has done something similar, then you're able to just
edit that, and that's saved a lot of work for me.''

Eric develops packages for analysts, and uses RCloud ``more for sharing code with
other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code... I write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets [and display the results].'' He also uses RCloud for
describing and debugging data sources.

@@

Automatic source control proved helpful. Lilo says, ``Instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

@@

\subsection{Discovering others' work}
Many users find the search function valuable. Wendy says, ``The fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''
But other users prefer to browse just the notebooks of experts they
know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
through, because I know that the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''
More selective ways to search will be needed as the number of notebooks
grows further. We mention possible approaches in Section~\ref{sec:discussion}.

@@

Integrating R helps Allison's application: ``Having an open session where I can run R commands or
functions without having to invoke an API or send a request and then
wait for the response is extremely helpful in writing the application.''

@@

Although most analysts appreciated RCloud's sharing features,
it was not a popular tool for exploratory data analysis.
All analysts with R experience prefered to do analysis in another
tool, then paste code into RCloud for sharing.
It appears this is mainly due to familiarity.
Eric asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

\subsection{Cells versus the command line}
RCloud's notebook interface combines editable Markdown with a command line
interface.
The interface takes some learning. Lilo says,
\begin{quote}
%% It took me a couple of weeks of
%% looking at it to become comfortable because 
I saw the cells and it just threw me
for a total loop. I mean, it's a good idea because [...] I can run it all at
once if I want or I can break these into sections for either debugging or
staging purposes. I really like it now, but when I first saw it, [it was] very
confusing.
\end{quote}

@@

The cell structure also can be problematic if some cells take a long time to
execute. High variance is common: cells may take anywhere from seconds to
hours to run. In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''
To avoid this pitfall, some users write code to explicitly cache results.
% Coby ends up ``littering the notebook with little switches that comment out the
% slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.

### 2015-06-25 relatedwork.tex
fixups

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it can be traced
back to Knuth's literate programming~\cite{Knuth:1984:LP}.
In literate programming, a prose description of the behavior of
a program is ``weaved'' with its source code, yielding both an
executable program and a human-readable document.
A notebook represented as a collection of small executable cells,
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally implemented as part of
IPython) offers some notebook features, though it lacks a transparent
mechanism for multi-user sharing and deployment.
%
Further afield, {Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.
%
RStudio is highly acclaimed for its polished and powerful human interface.
Although RStudio offers publication of literate R programs as a free
service on its website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for other users to build off the work published, or even for the
original author to update new versions. In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.

\paragraph*{Provenance and versioning} As
mentioned, one central issue in exploratory analysis is that
problems may change quickly, often while in the course of
developing a solution. As a result, systems need to provide adequate
support for tracking \emph{changes} of data analysis scripts. VisTrails
was one of the original systems for managing \emph{process provenance},
and it demonstrated the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

VisMashup~\cite{Santos:2009:VST} defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website
feature~\cite{Mates:2011:CSA} workflow upload and remote execution. In
our view, the impedance mismatch between a dataflow pipeline
specification and the power of a general-purpose language is too great
for the type of general exploratory work in data science teams. At the
expense of ease of use for non-programmers, RCloud tries to provide a
closer match for analysts accustomed to creating and executing R and
Python code, while retaining attractive properties like transparent
provenance tracking and interactive data visualization on the web.

\paragraph*{Web-based tools for sharing code snippets}
Quite a few tools were recently developed for quickly
sharing small programs on the web, including
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
share Javascript programs, which means that deployment happens
automatically through the web browser. If Javascript eventually becomes
the lingua franca of exploratory data analysis, we foresee
building a simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not a realistic assumption at present, making these stand-alone
solutions unsuitable for our purpose. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: in other
words, it is possible to generate a plot.ly visualization from inside
another program. Although this is an intriguing idea, it
nevertheless creates a disconnection between the analysis and the
resulting visualization. In RCloud, we wanted to ensure that every
visualization is transparently linked to the source code that
generates it.

\paragraph*{Needs of data analysts}
Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. Kandel et al also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is difficult and important, and that sharing
and versioning of data sources and artifacts is hindered by current
technology. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their study highlights the opportunity to contribute better
ways of supporting collaboration and sharing by data analysis teams.

An earlier study by Kandel et al argues that data wrangling
(cleaning, parsing and transformation)is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. We view this
as attacking a different semantic level than ours, but also
showing the need for an environment that enables better sharing
of the knowledge, tools and processes to do this. Anecdotally
we find much frustration among practitioners that this knowledge
is difficult to find and often is not recorded or available in a
reusable form even within the same organization.

@@

The need for integrating statistics and visualization has been
highlighted in previous studies and is widely understood by
various technical communities \cite{Perer:2008:ISA}
Lucas and Roth were early advocates of combining
data exploration with presentation and publication \cite{Lucas:1996:EIV}.

>>>>

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it can be traced
to Knuth's literate programming~\cite{Knuth:1984:LP}.
In literate programming, a prose description of the behavior of
a program is ``weaved'' with its source code, yielding both an
executable program and a human-readable document.
A notebook represented as a collection of small executable cells
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally part of
IPython) offers some notebook features, though it lacks transparent
multi-user sharing and deployment.
RStudio is highly acclaimed for its polished and powerful human interface.
Although RStudio offers publication of literate R programs as a free
service on its website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for others to build off the work published, or even for the
original author to update new versions. In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.
%
% This may be something we can delete -- scn
Further afield, {Electronic Lab Notebooks} organize
and share data from scientific lab experiments\cite{Rubacha:2011:ELN}.
% In a sense, we hope to adapt and extend this concept to the work of visual analytics teams.
%

\paragraph*{Provenance and versioning} As
mentioned, one central issue in exploratory analysis is that
problems may change quickly, often while in the course of
developing a solution. As a result, systems need to provide adequate
support for tracking \emph{changes} of data analysis scripts. VisTrails
was an early system for managing \emph{process provenance},
that demonstrated the value of capturing aspects of the processes
that surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

VisMashup~\cite{Santos:2009:VST} defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website
featuring workflow upload and remote execution~\cite{Mates:2011:CSA}.
In our view, the impedance mismatch between dataflow pipeline
specifications and powerful general-purpose languages is too great
for exploratory work by data science teams. At the
expense of ease of use for non-programmers, RCloud provides
a closer match for analysts familiar with R and Python,
while retaining attractive properties like transparent
provenance tracking and interactive data visualization on the web.

\paragraph*{Web-based tools for sharing code snippets}
Quite a few tools were recently developed for quickly
sharing small programs on the web, including
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
share Javascript programs, which means that deployment happens
automatically through the web browser. If Javascript becomes
the lingua franca of exploratory data analysis, we foresee
building a simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not realistic today, making these stand-alone
solutions unsuitable for our purpose. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: 
it is possible to generate a plot.ly visualization within
another program. Although this is an intriguing idea, it
nevertheless creates a disconnection between the analysis and the
resulting visualization. In RCloud, we wanted to ensure that every
visualization is transparently linked to the source code that
generates it.

\paragraph*{Needs of data analysts}
Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome.
% RCloud seeks to reduce this mismatch.
Kandel et al also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is difficult and important, and that sharing
and versioning of data sources and artifacts is hindered by current
technology. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.''
% Their study highlights the opportunity to contribute better
% ways of supporting collaboration and sharing by data analysis teams.

An earlier study by Kandel et al argues that data wrangling
(cleaning, parsing and transformation)is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}.
% We view this as attacking a different semantic level than ours, but also showing
Although attacking a different semantic level from ours, it shows
the need for an environment that enables better sharing
of the knowledge, tools and processes to do this. Anecdotally
we find much frustration among practitioners that this knowledge
is difficult to find and often is not recorded or available in a
reusable form even within the same organization.

@@

% The need for integrating statistics and visualization has been
% highlighted in previous studies and is widely understood by
% various technical communities \cite{Perer:2008:ISA}
% Lucas and Roth were early advocates of combining
% data exploration with presentation and publication \cite{Lucas:1996:EIV}.
